Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})

### 2017-09-05 doc/turtles_statistical_learning.tex
updated 7cc data. stuart and ken comments on manuscript.

%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
\title{Supervised learning approaches to classifying morphological variation: assessing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
\author[1]{Peter D Smits}%\thanks{psmits@uchicago.edu}}
\author[1,2]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[3]{James F Parham}%\thanks{jparham@fullerton.edu}}
\author[4]{Bryan Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
\affil[1]{Committee on Evolutionary Biology, University of Chicago}
\affil[2]{Integrative Research Center, Field Museum of Natural History}
\affil[3]{Department of Geological Sciences, California State University -- Fullerton}
\affil[4]{Section of Research and Collections, North Carolina Museum of Sciences}

\begin{document}
\maketitle
\noindent{\textbf{Corresponding author:} Peter D Smits, Committee on Evolutionary Biology, University of Chicago, 1025 E. 57th Street, Culver Hall 402, Chicago, IL, 60637, USA; E-mail: \href{mailto:psmits@uchicago.edu}{psmits@uchicago.edu}}

@@

\begin{abstract}
  We investigate morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}). We collected landmark-based morphometric data from 354 adult \textit{E. marmorata/``pallida''} museum specimens. We assigned a classification to each specimen for six different binning schemes using geographic occurrence data recorded in museum collection archives. We used multiple supervised machine learning methods to compare different classification hypotheses and asked whether it is possible to determine which among this set of classification hypotheses is best. 
  In addition, we applied the above approach to two other datasets: two well-supported subspecies of \textit{Trachemys scripta} and a set of seven unambiguously distinct species closely related to \textit{E. marmorata}. The results of this study indicate that there is no clear grouping of \textit{E. marmorata/``pallida''} based on plastron shape. In contrast, analysis of the other datasets resulted in near perfect classifications, which demonstrates that application of the methods to plastron shape data can recover correct results in similar situations. Explanations for the lack of grouping in \textit{E. marmorata} include that possibility that genetic differentiation is not associated with plastron shape variation in this species complex and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. A reconsideration of the methods used to delimit \textit{E. ``pallida,''} the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to differentiate other emydid species and subspecies suggest that the non-diagnosability of \textit{E. ``pallida''} stems from it not being a distinct species.
\end{abstract}

@@

In contrast to those genetically-diagnosed species, the majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone. This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,Hunt2008,VanBocxlaer2013}. It also has serious implications for our records of modern biodiversity: for many museum specimens of extant taxa (e.g. those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.

These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species. This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations. Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009}. Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}.

Here, we investigate the morphometric identification of cryptic species using machine learning approaches. In particular, we ask whether it is possible to determine which among a set of classification hypotheses is best and examine the implications of the results for a recently proposed set of cryptic turtle species.

@@

Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013}. Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes. They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results. Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.

Here, we used multiple supervised machine learning methods to compare different classification hypotheses. Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications. Although machine learning methods such as neural networks have been applied to studying shape variation \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}, including in the context of automated taxon identification and classification of groups, the number of cases remains limited. In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.

@@

More recently, the phylogeography of \textit{E. marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}. Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E. marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade. Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E. marmorata}, although these datasets differed from that of \citet{Spinks2005} in the location of the break point. All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E. marmorata} and \emph{E. pallida}), and a possible third based on populations in Baja California. However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}. Given that these characters are variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E. marmorata}.

In this study, we attempt to estimate the best classification scheme of \textit{E. marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the past divisions based on other morphological features or molecular data. We are particularly interested in whether any aspect of plastron shape can be used to reliably diagnose the proposed species of \citet{Spinks2014}, and if so, the nature of that shape variation.

\uppercase{Why did we analyze plastron shape? Peter doesn't have a good answer.}

Because of unclear geographic boundaries between subgroups of \textit{E. marmorata}, we compare multiple hypotheses of morphologically-- and molecularly--based classification. We hypothesize that if morphological variation corresponds to class assignment, it should be possible to determine the best classification hypothesis of \textit{E. marmorata} from amongst multiple candidate hypotheses. However, if morphological variation does not correspond to any of the standing hypothesis, then supervised learning model generalization performance will be poor.

\section{Materials and Methods}
\subsection{Specimens, sampling, morphometrics}
Three different landmark-based morphometric datasets describing plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T. scripta} specimens from the two main subspecies (\textit{T. scripta elgans} and \textit{T. scripta scripta}); and 3) \textit{E. marmorata} specimens from across the species' geographic range. The first two datasets are intended to serve as a basic test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape. Dataset 1 should be relatively easy, given the high level of plastron shape disparty that exists among these species \citep{Angieczyk2011}, whereas dataset 2 should be a case that is more comparable to our \textit{E. marmorata} dataset. We chose to focus on adults in all three cases because significant changes in plastron shape occur over the course of ontogeny in \textit{E. marmorata} and other emydines \citep{Angielczyk2013a}.
% three turtle datasets
%   seven species
%   Trachemys species
%   Emys complex
% 26 landmarks

@@

The final dataset dataset included 354 adult \textit{E. marmorata} museum specimens; representing a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}. We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives. When precise latitude and longitude information were not available we estimated them from locality information. Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies. Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented some hypothesis with two schemes for a total of six different schemes. These schemes differed based on where geographic boundaries were assigned. This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes. Sex information was known only for a subset of the total dataset and was not included as a predictor of classification. However, we did use sex information to determine if observations cluster by sex or not. The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}. % CHANGE

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmark used in this study. Anterior is towards the top of the figure.}
  \label{fig:plastra}
\end{figure}

Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig. \ref{fig:plastra}). Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron. Twelve of the landmarks were symmetrical across the axis of symmetry. Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e. midline) prior to analysis and used the average position of each symmetrical pair. In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair. We conducted all subsequent analyses on the resulting ``half'' plastra. We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which, we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.

\subsection{Biasing effects}
% digitization
We estimated the possible effect of digitization error \citep{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids. Ten randomly selected \textit{E. marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate. These 50 landmark configurations were then Procrustes superimposed. A range of four Procrustes distances were then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.

@@

Turtles are well known to demonstrate strong sexual dimorphism in plastron shape and morphology (e.g. the presence of a plastral concavity) CITATION. To test for biases resulting from sexual dimorphism in our \textit{E. marmorata} data set, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled. Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data.

\subsection{Supervised learning approaches}
Instead of relying on a single supervised learning method, we chose to use an ensemble of multiple approaches so that the congruence between the them could be used as a means of ``support'' for one conclusion or another. The supervised learning methods used here are described in Table \ref{tab:methods}. Each of these methods makes different assumptions, treat data differently, and can produce different qualities of classification results depending on the nature of the data \citep{Hastie2009}. 
% use multiple methods
%   multinomial logistic regression (nnet)
%   linear discriminate analysis (MASS)
%   penalized discriminate analysis (mda)
%   neural network (nnet)
%   random forest (randomForest)
% each method has different assumptions and treat data differently
%   all assume that predictors have additive effect (e.g. independent)

@@

% predictors/features/covariates
%   1:25 PCs
%   size
%   size X PC1
% analogy to PCA regression
The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1. Additional interaction terms were not considered because of model complexity/sample size concerns. Size and the interaction between size and PC 1 were included as predictors t account for known ontogenetic variation in plastron shape \citep{Angieczyk2013a} as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}. These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below).

Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression. PCA regression takes advantage of dimensionality reduction and orthogonal PCs aspects of PCA to improve regression fit \citep{Hastie2009}. Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.

% training and testing paradigm
%   training dataset
%     for multiple models with between 3 and 28 predictors
%     5-fold cross validation of each model
%     best model had greatest mean AUC value
%     selected model is most parsimonious model within 1 SD of ROC of best model
%   testing dataset
%     estimate class of out-of-sample testing dataset
%     compare average AUC across models for each scheme
% caret package for R
We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \citep{Hastie2009,Kuhn2013}. Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias. With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data is split into a training dataset and a testing dataset. The former is used for fitting the model where as the later is used for measuring model performance, a process called model generalization. In this analysis, we used 80\% of samples as the training set while the remaining 20\% were used as the testing set.

@@

For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation. Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \citep{Hastie2009}. In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted. This is repeated for all combinations of blocks. Within each round, the predictive performance metrics are averaged across all folds. Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation. This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}.

For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation. The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}. This rule is applied to ameliorate the chance of selecting an overly complex model.

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[b]{0.4\textwidth}
    \caption{}
    \includegraphics[width = \textwidth]{figure/cc7_pc_graph}
  \end{subfigure}
  \begin{subfigure}[b]{0.4\textwidth}
    \caption{}
    \includegraphics[width = \textwidth]{figure/tra_pc_graph}
  \end{subfigure}
  \caption{Scatterplots of the first two axes from the PCA results for both the seven species (a) and \textit{Trachemys} (b) datasets. Point color corresponds to the species within each dataset. Point size is proportional to individual centroid size. In parentheses next to the axis labels are the percent of total variation accounted for by that axis.}
  \label{fig:other_pca}
\end{figure}

The results of the PCA of plastron shape in the \textit{Emys marmorata} dataset show no clear connection between plastron shape and any of the six proposed classification schemes (Fig. \ref{fig:emys_pca}). The first PC axis of shape variation appears to be very structured by differences in individual centroid size (Fig. \ref{fig:emys_pca}); this was the motivation for including centroid size and its interaction with PC1 as predictors in all of the supervised learning models.

>>>>

%\title{How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})}
\title{Ensemble approaches for estimating congruence between species delimitation and morphological variation: comparing taxonomic hypotheses for the Pacific Pond Turtle (\textit{Emys marmorata})}
\author[1]{Peter D Smits}%\thanks{peterdavidsmits@gmail.com}}
\author[2,3]{Kenneth D Angielczyk}%\thanks{kangielczyk@fieldmuseum.org}}
\author[4]{James F Parham}%\thanks{jparham@fullerton.edu}}
\author[5]{Bryan L Stuart}%\thanks{bryan.stuart@naturalsciences.org}}
\affil[1]{Department of Integrative Biology, University of California -- Berkeley}
\affil[2]{Committee on Evolutionary Biology, University of Chicago}
\affil[3]{Integrative Research Center, Field Museum of Natural History}
\affil[4]{Department of Geological Sciences, California State University -- Fullerton}
\affil[5]{Section of Research and Collections, North Carolina Museum of Natural Sciences}

\begin{document}
\maketitle
\noindent{\textbf{Corresponding author:} Peter D Smits, Department of Integrative Biology, University of California -- Berkeley, 3040 Valley Life Sciences Building \#5151, Berkeley, CA, 94720, USA; E-mail: \href{mailto:peterdavidsmits@gmail.com}{peterdavidsmits@gmail.com}}

@@

\begin{abstract}
%  We investigated the morphometric identification of cryptic species using machine learning approaches by examining their implications for a recently proposed cryptic turtle species (\textit{Emys pallida}). We collected landmark-based morphometric data from 366 adult \textit{E. marmorata/``E. pallida"} museum specimens. We assigned a classification to each specimen for six different binning schemes based on geographic occurrence data recorded in museum collection archives. We used an ensemble of supervised machine learning approaches to determine which classification hypothesis was best supported by the data. In addition, we applied the same approach to two clear-cut examples, one consisting of seven unambiguously distinct species closely related to \textit{E. marmorata}, and the other consisting of two subspecies of \textit{Trachemys scripta}. Our results indicate that there is no clear grouping of \textit{E. marmorata/``E. pallida"} based on plastron shape. In contrast, the analyses of the clear-cut examples produced near perfect classifications, demonstrating that the methods can recover correct results when an appropriate signal exists. Explanations for the lack of grouping in \textit{E. marmorata} include the possibility that genetic differentiation is not associated with plastron shape variation below the species level and/or that local selective pressures (e.g., from hydrological regime) overwhelm morphological differentiation. A reconsideration of the methods used to delimit \textit{``E. pallida"}, the lack of barriers to gene flow, the strong evidence for widespread admixture between lineages, and the fact that plastron shape can be used to delineate other emydine species and sub-species suggest that its lack of diagnosability most likely reflects the non-distinctiveness of this proposed taxon.
\end{abstract}

@@

In contrast to those genetically-diagnosed species, the majority of extant taxa, and almost all extinct taxa, are delimited by morphology alone. This disjunction complicates interpretations of variation and diversity in deep time, as apparent morphological stasis may not reflect the true underlying diversity \citep{Eldredge1972,Gould1977a,VanBocxlaer2013}. It also has serious implications for our records of modern biodiversity: for many museum specimens of extant taxa (e.g. those preserved in formalin), it is difficult to acquire the genetic data needed for non-morphological species delimitation methods.

These considerations have sparked interest in whether geometric morphometric analyses can capture fine-scale variation that can be used for identifying cryptic species. This would make the task of identifying and maintaining endangered or conserved groups much easier and could contribute to improved classifications of extinct taxa and populations. Most such studies focus on using morphometrics to discover differences between taxa that were identified by other means \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Demandt2009,Markolf2013,Fruciano2016}. Additionally, there has been work on automated taxon identification and classification of taxa into groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011,Vitek2017}.

Here, we investigate the morphometric identification of cryptic species using machine learning approaches. We use an ensemble learning approach where multiple methods are used in order to look for consensus amoung their results. We test our approach on three datasets: plastron shape of seven species of closely related turtles, plastron shape of two species of closely related turtles, and plastron shape of the \textit{Emys marmorata} species complex. In particular, we ask whether it is possible to determine which among a set of classification hypotheses best aligns with the observed morphology, and examine the implications of our results for the \textit{E. marmorata} complex.

@@

Geometric morphometric approaches to identifying differences in morphological variation between different classes, including cryptic species, mostly have used methods like linear discriminate analysis and canonical variates analysis \citep{Polly2003,Zelditch2004,Gaubert2005b,Gunduz2007,Polly2007a,Francoy2009,Sztencel-Jabonka2009,MitrovskiBogdanovic2013,Dillard2017}. Because of their similarity to multivariate approaches like principal components analysis (PCA), these methods are comparatively straightforward ways of understanding the differences in morphology between classes. They also benefit from producing results that can be easily visualized, which aids in the interpretation and presentation of data and results. Most previous morphometric studies did not assess which amongst a set of alternative classification hypotheses was optimal. For example, studies such as those of \citet{Caumul2005a} and \citet{Polly2007a} focused on comparing different aspects of morphology and their fidelity to a classification scheme instead of comparing the fidelity of one aspect of morphology to multiple classification schemes. In this context, the study of \citet{Cardini2009a} is noteworthy because they compared morphological variation in marmots at the population, regional, and species level and determined the fidelity of shape to divisions at each of these levels.

Here, we used an ensemble of supervised machine learning methods to compare the congruence of the morphological data to different classification hypotheses. Each of these methods provide different advantages for understanding how to classify taxa, as well as the accuracy of the resulting classifications. 
Machine learning methods have been combined with geometric morphometric data to study shape variation in a variety of contexts, including automated taxon identification and classificaiton of groups \citep{Baylac2003,Dobigny2003,MacLeod2007,VandenBrink2011}. % MORE
%Although machine learning methods such as neural networks have been applied to studying shape variation, including in the context of automated taxon identification and classification of groups, the number of cases remains limited. 
In the current study, we not only consider pure classification accuracy but also use a statistic of classification strength that reflects the rate at which taxa are both accurately and inaccurately classified.

@@

More recently, the phylogeography of \textit{E. marmorata} and the possibility of cryptic diversity was investigated using molecular data \citep{Spinks2005,Spinks2010,Spinks2014}. Based on mitochondrial DNA, \citet{Spinks2005} recognized four subclades within \textit{E. marmorata}, a northern clade, a San Joaquin Valley clade, a Santa Barbara clade, and a southern clade. Analyses with nuclear DNA \citep{Spinks2010} and single-nucleotide polymorphism (SNP) data suggest a primarily north--south division in \textit{E. marmorata}, although these datasets differed from that of mitocondrial-based results of \citet{Spinks2005} in the location of the break point \citep{Spinks2014}. All three studies discussed the potential taxonomic implications of their results, with \citet{Spinks2014} going so far as to strongly advocate for the recognition of at least two species (\emph{E. marmorata} and \emph{E. pallida}), and a possible third based on populations in Baja California. However, they did not discuss in detail the morphological characters that would help to diagnose these species beyond those specified by \citet{Seeliger1945}. Given that these characters are variable within the proposed species, and that \citet{Holland1992} described shell shape variation that might be consistent with this taxonomy, a geometric morphometric analysis of shell shape might provide a reliable way to diagnose groups (whether species or subspecies) within \textit{E. marmorata}.

In this study, we attempt to estimate the best classification scheme of \textit{E. marmorata} based on variation in plastron (ventral shell) shape in order to determine whether this character is consistent with any of the proposed taxonomies of the \textit{E. marmorata} complex.

We choose to analyze plastron shape for multiple reasons. First, it is very easy to collect geometric morphometric data on plastron shape from two-dimensional pictures as the structure is virtually flat. This approach allows both museum specimens and individuals in the field to be analyzed together. Second, previous work has suggested that there are strong differences in plastron shape among traditionally recognized emydine species \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}. Finally, due to this previous study a large dataset was readily avaliable.

In the case of the \textit{E. marmorata} species complex, we hypothesize that if one or more of the proposed classification schemes are consistent with the morphological data then our ensemble approach fit to those hypotheses will have higher out-of-sample predictive performance than the more inconsistent hypotheses. However, if all of the classification schemes lead to equal out-of-sample predictive performance then we would conclude that the molecular based hypotheses are inconsistent with whatever information is present in the morphological data. Because of unclear geographic boundaries between subgroups of \textit{E. marmorata}, we compare multiple permutations of the \citep{Spinks2010} and \citet{Spinks2014} hypotheses.

\section{Materials and Methods}
\subsection{Specimens, sampling, morphometrics}
Three different geometric morphometric datasets describing turtle plastron variation were assembled for this analysis: 1) specimens from seven distinct emydine species; 2) \textit{T. scripta} specimens from the two main subspecies (\textit{T. scripta elegans} and \textit{T. scripta scripta}); and 3) \textit{E. marmorata} specimens from across the species' geographic range. The first two datasets are intended to serve as a test of whether machine learning techniques can differentiate species-level groupings of emydine turtles using plastron shape. We expect that the first case represents a low complexity dataset because of the high level of plastron shape disparity that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity and more analogous to the \textit{E. marmorata} example. We predict that the \textit{E. marmorata} dataset should be of the highest complexity and our greatest challenge given \citep{Holland1992} finding that only very subtle differences existed between geographically-distinct populations.
%We expect the first dataset represents a low-complexity dataset given the high level of plastron shape disparty that exists among these species \citep{Angielczyk2011}, whereas the second dataset should be relatively higher in complexity than the first. Ultimately, we expect that the third dataset to be of the highest complexity and our biggest analytical challenge. For all datasets we chose to focus on adults because significant changes in plastron shape occur over the course of ontogeny in \textit{E. marmorata} and other emydines \citep{Angielczyk2013a}.
% three turtle datasets
%   seven species
%   Trachemys species
%   Emys complex
% 26 landmarks

@@

The final dataset is of 366 adult \textit{E. marmorata} museum specimens. These speciemens represent a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}. Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample the individuals in their studies. Instead, our specimen classifications were based solely on the geographic information and not explicit assignment using molecular data. For each taxonomic hypothesis, specimens were assigned to one of the possible classes based on geographic occurrence data recorded in museum collection archives. In cases where precise latitude and longitude information were not available we estimated for them from other locality information. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with multiple possible realizations representing the classification uncertainty for specimens present at the geographic boundaries. The taxonomic hypotheses and sub-hypotheses for the \textit{E. marorata} used here are presented in Table \ref{tab:hypotheses}.

Sex was known only for a subset of the total dataset and was not included as a predictor of classification. Instead, we estimated the degree by which specimens cluster morphologically by sex in order to determine how much of a potential biasing factor sexual dimorphism could be for our analysis of the \textit{E. marmorata} species complex (see below).
%The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}. % CHANGE

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/plastra}
  \caption{Depiction of general plastral shape of \textit{E. marmorata} and position of the 19 landmarks used in this study. Anterior is towards the top of the figure.}
  \label{fig:plastra}
\end{figure}

\begin{table}
  \centering
  \caption{Table of species delimitation hypotheses for \textit{E. marmorata}}
  \begin{tabular}{l l l }
    \hline
    Abbreviation & Number of classes & citation \\
    \hline
    SP10.1 & 3 & \citet{Spinks2010} \\
    SP10.2 & 3 & \citet{Spinks2010} \\
    SP10.3 & 4 & \citet{Spinks2010} \\
    SP14.1 & 2 & \citet{Spinks2014} \\
    SP14.2 & 4 & \citet{Spinks2014} \\
    \hline
  \end{tabular}
  \label{tab:hypotheses}
\end{table}

Following previous work on plastron shape \citep{Angielczyk2007,Angielczyk2011,Angielczyk2013a}, we used TpsDig 2.04 \citep{Rohlf2005} to digitize 19 two-dimensional landmarks (Fig. \ref{fig:plastra}). Seventeen of the landmarks are at the endpoints or intersection of the keratinous plastral scutes that cover the plastron. Twelve of the landmarks were symmetrical across the axis of symmetry. Because damage prevented the digitization of all the symmetric landmarks in some specimens, we reflected landmarks across the axis of symmetry (i.e. midline) prior to analysis and used the average position of each symmetrical pair. In cases where damage or incompleteness prevented symmetric landmarks from being determined, we used only the single member of the pair. We conducted all subsequent analyses on the resulting ``half'' plastra. We superimposed the plastral landmark configurations using generalized Procrustes analysis \citep{Dryden1998a}, after which we calculated the principal components (PC) of shape using the \texttt{shapes} package for R \citep{R2016,Dryden2013}.

\subsection{Biasing effects}
% digitization
We estimated the possible effect of digitization error \citep{Arnqvist1998,Cramon2007,Munoz-MunozF.2010} on our results by comparing within (replicated) specimen Procrustes distances to the distances between classification scheme centroids. Ten randomly selected \textit{E. marmorata} specimens were each digitized four times, with the original set of digitized coordinates serving as a fifth replicate. These 50 landmark configurations were then Procrustes superimposed. A range of four Procrustes distances was then calculated as the average of the pairwise distances between each of the replicate configurations of a given specimen.

@@

\textit{Emys marmorata} is known to display sexual dimorhpism in plastron shape, particulary the presence of a plastra concavity in males \citep{Seeliger1945}. To test for biases resulting from sexual dimorphism in our \textit{E. marmorata} dataset, we used a simple permutation test to determine if the distance between the mean female and male shapes is greater than expected when the sex labels are randomly shuffled. Because not all of our specimens have sex identifications associated with them, this analysis was done using a subset of the data (257 of 360).

\subsection{Supervised learning approaches}
Instead of relying on a single supervised learning method, we chose to use an ensemble approach where multiple model types are used in concert so that any congruence between the them increases our support for that conclusion over another \citep{Hastie2009}. The supervised learning methods used here are named in Table \ref{tab:methods}. Each of these methods makes different assumptions, treats data differently, and can produce different classification results depending on the nature of the data \citep{Hastie2009}. For example, multinomial logistic regression is a type of generalized linear model while random forest is an ensemble approach where multiple decision trees are fit to subsets of the full data and then averaged.
% use multiple methods
%   multinomial logistic regression (nnet)
%   linear discriminate analysis (MASS)
%   penalized discriminate analysis (mda)
%   neural network (nnet)
%   random forest (randomForest)
% each method has different assumptions and treat data differently
%   all assume that predictors have additive effect (e.g. independent)

@@

% predictors/features/covariates
%   1:25 PCs
%   size
%   size X PC1
% analogy to PCA regression
The maximum set of possible predictors or features used for any model of our dataset is comprised of the first 25 principal components (PCs), scaled centroid size, and the interaction between scaled centroid size and PC 1. Additional interaction terms were not considered because of model complexity/sample size concerns. Size and the interaction between size and PC 1 were included as predictors to account for known ontogenetic variation in plastron shape \citep{Angielczyk2013a} as well as potential size differences between classes, even if this is unlikely \citep{Seeliger1945,Holland1992}. These data constitute a ``maximum set'' because the best or selected models based on five-fold cross-validation need not, and likely will not, include all predictors possible (see below). Because our supervised learning models use PCs as predictors, this approach is in many ways analogous to PCA regression. PCA regression takes advantage of reduction and orthogonality PCs to improve regression fit \citep{Hastie2009}. Because the PCs of shape are by definition orthogonal, they can easily serve as independent predictors or features of class membership without fear of collinearity.

% training and testing paradigm
%   training dataset
%     for multiple models with between 3 and 28 predictors
%     5-fold cross validation of each model
%     best model had greatest mean AUC value
%     selected model is most parsimonious model within 1 SD of ROC of best model
%   testing dataset
%     estimate class of out-of-sample testing dataset
%     compare average AUC across models for each scheme
% caret package for R
We adopted a training and testing paradigm for selecting parsimonious models and estimating their overall error rates \citep{Hastie2009,Kuhn2013}. Within-sample model performance is inherently biased upwards, so model evaluation requires overcoming this bias. With very large sample sizes, as in this study, part of the sample can be used as the ``training set'' and the remainder acts as the ``testing set.'' In this approach, following all cleaning and vetting, the data are split into a training dataset and a testing dataset. The former is used for fitting the model whereas the later is used for measuring model performance, a process called model generalization. In this analysis, we used 80\% of samples as the training set while the remaining 20\% were used as the testing set.

@@

For a given supervised learning method, we compared the fit of 27 models as the average AUC from 10 rounds of five-fold cross-validation. Cross-validation is an approach for estimating the average out-of-sample predictive error of a model by simulating out-of-sample data from the training dataset itself \citep{Hastie2009}. In a single round of \(k\)-fold cross-validation, the training data are divided into \(k\) blocks where the model is fit to \(k - 1\) blocks and the values of the \(k\)th block are predicted. This is repeated for all combinations of blocks. Within each round, the predictive performance metrics are averaged across all folds. Finally, the predictive performance metric is the averaged across all rounds of \(k\)-fold cross-validation. This process was implemented using the R package \texttt{caret} \citep{KuhnMAN2013}. For a given supervised learning method, the ``best'' trained model is that with the highest mean AUC as estimated from five-fold cross-validation. The selected or final model, however, is the next most parsimonious model that is within one standard error of the best model; this is a variant on the ``one-standard error'' rule from \citet{Hastie2009}. This rule is applied to ameliorate the chance of selecting an overly complex model.

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[b]{0.7\textwidth}
    \caption{}
    \includegraphics[width = \textwidth]{figure/cc7_pc_graph}
  \end{subfigure}

  \begin{subfigure}[b]{0.7\textwidth}
    \caption{}
    \includegraphics[width = \textwidth]{figure/tra_pc_graph}
  \end{subfigure}
  \caption{Two scatterplots of morphological differences from two of the three datasets analyzed in this study. (a) Scatterplot of the first two PCA axies from the landmarks from the seven different species dataset, and (b) the first two axes of variation from two subspecies of \textit{Trachemys} dataset. Point colors correspond to the categories within each dataset while point size is proportional to individual centroid size. In parentheses next to the axis labels are the percent of total variation accounted for by that axis. For both datasets there are clear distinctions between the different categories.}
  \label{fig:other_pca}
\end{figure}
